\begingroup
\def\hd{\begin{tabular}{lll}
          \textbf{Document Number:} & {\larger\docno}                 \\
          \textbf{Date:}            & \reldate                        \\
          \textbf{Audience:}        & Library Evolution Working Group \\
          \textbf{Author:}          & Casey Carter                    \\
          \textbf{Reply to:}        & casey@carter.net
          \end{tabular}
}
\newlength{\hdwidth}
\settowidth{\hdwidth}{\hd}
\hfill\begin{minipage}{\hdwidth}\hd\end{minipage}
\endgroup

\vspace{0.5cm}
\begin{center}
\textbf{\Huge \doctitle}
\end{center}

\vspace{0.5cm}

\chapter{Overview}

The coroutines TS has introduced a language capability that allows functions to be suspended and later resumed. One of the key applications for this new feature is to make it easier to write asynchronous code. However, the coroutines TS itself does not include any concrete coroutine types that directly support writing asynchronous code.

This paper proposes adding a new type, \tcode{std::lazy<T>}, to the standard library to enable creation and composition of coroutines representing asynchronous computation.

\begin{note}
SG1 wants to reserve the name \tcode{std::task} for something else. SG1 requested to pick a different name. LEWG chose \tcode{lazy}.
\end{note}

\begin{codeblock}
#include <lazy>
#include <string>

struct record {
  int id;
  std::string name;
  std::string description;
};

std::lazy<record> load_record(int id);
std::lazy<> save_record(record r);

std::lazy<void> modify_record() {
  record r = co_await load_record(123);
  r.description = "Look, ma, no blocking!";
  co_await save_record(std::move(r));
}
\end{codeblock}

The interface of lazy is intentionally minimal and designed for efficiency. In fact, the only operation you can do with the lazy is to await on it.

\begin{codeblock}
template <class T>
class [[nodiscard]] lazy {
public:
  lazy(lazy&& that) noexcept;
  ~lazy();
  unspecified operator co_await(); // \expos
};
\end{codeblock}

While such small interface may seem unusual at first, subsequent sections will clarify the rationale for this design.

\section{Why not use futures with \tcode{future.then}?}

The \tcode{std::future} type is inherently inefficient and cannot be used for efficient composition of asynchronous operations.
The unavoidable overhead of futures is due to:
\begin{itemize}
\item allocation/deallocation of the shared state object,
\item atomic increment/decrement for managing the lifetime of the shared state object,
\item synchronization between setting of the result and getting the result, and
\item (with \tcode{.then}) scheduling overhead of starting execution of subscribers to \tcode{.then}.
\end{itemize}

Consider the following example:
\begin{codeblock}
lazy<int> coro() {
  int result = 0;
  while (int v = co_await async_read())
    result += v;
  co_return result;
}
\end{codeblock}
where \tcode{async_read()} is some asynchronous operation that takes, say, 4ns to perform.
We would like to factor out the logic into two coroutines:
\begin{codeblock}
lazy<int> subtask() {
  co_return co_await async_read();
}

lazy<int> coro_refactored() {
  int result = 0;
  while (int v = co_await subtask())
    result += v;
  co_return result;
}
\end{codeblock}

Breaking a single \tcode{co_await} into its own function may seem silly, but it allows us to measure the overhead of composition of tasks. With the proposed \tcode{lazy}, our per-operation cost grew from 4ns to 6ns and did not incur any heap allocations. Moreover, this overhead of 2ns is not inherent to \tcode{lazy} and we anticipate that with improved coroutine optimization technology we will be able to drive the overhead close to zero.
To estimate the cost of composition with \tcode{std::future}, we used the following code:
\begin{codeblock}
int fut_test() {
  int count = 1'000'000;
  int result = 0;

  while (count > 0) {
    promise p;
    auto fut = p.get_future();
    p.set_value(count--);
    result += fut.get();
  }
  return result;
}
\end{codeblock}
As measured on the same system (Linux, clang 6.0, libc++), we get \emph{133ns} per operation! Here is the visual illustration.

\begin{tabular}{rl}
        op cost: & ** \\
  lazy overhead: & * \\
future overhead: & ***************************************************************** \\
\end{tabular}

Being able to break apart bigger functions into a set of smaller ones and being able to compose software by putting together small pieces is fundamental requirement for a good software engineering since the 60s. The overhead of \tcode{std::future} and types similar in behavior makes them unsuitable as coroutine types.

\subsubsection{Removing \tcode{future} overhead: Memory Allocation}

Consider the only operation that is available on a \tcode{lazy}, namely, awaiting on it.
\begin{codeblock}
lazy<X> g();
lazy<Y> f() {
  // ...
  X x = co_await g();
  // ...
}
\end{codeblock}
The caller coroutine \tcode{f} owns the lazy object for \tcode{g} that is created and destroyed at the end of the full expression containing \tcode{co_await}. This allows the compiler to determine the lifetime of the coroutine and apply Heap Allocation eLision Optimization~\cite{p0981r0}, which allocates a coroutine's state as if it were a local variable in its caller.

\subsubsection{Removing \tcode{future} overhead: Reference counting}

The coroutine state is not shared. The \tcode{lazy} type only allows moving pointer to a coroutine from one \tcode{lazy} object to another. Lifetime of a coroutine is linked to its \tcode{lazy} object, and the \tcode{lazy} object's destructor destroys the coroutine, thus, no reference counting is required.

In a later section about cancellation we will cover the implications of this design decision.

\subsubsection{Removing \tcode{future} overhead: Set/Get synchronization}

The \tcode{lazy} coroutine always starts suspended. This allows not only to avoid synchronization when attaching a continuation, but also enables solving via composition how and where coroutine needs to get executed and allows to implement advanced execution strategies like continuation stealing.

\subsubsection{Removing \tcode{future} overhead: Scheduling overhead}
Consider the following code fragment:
\begin{codeblock}
int result = 0;
while (int v = co_await async_read())
  result += v;
\end{codeblock}
Let's say that \tcode{async_read} returns a \tcode{future}. That \tcode{future} cannot resume directly the coroutine that is awaiting on it as it will, in effect, transform the loop into unbounded recursion.

On the other hand, coroutines have built-in support for symmetric coroutine transfer~\cite{p0913r1}. Since \tcode{lazy} object can only be created by a coroutine and the only way to get the result from a coroutine is by awaiting on it from another coroutine, the transfer of control from completing coroutine to awaiting coroutine is done in symmetric fashion, thus eliminating the need for extra scheduling interactions.

\subsection{Destruction and cancellation}
Note that the \tcode{lazy} type unconditionally destroys the coroutine in its destructor. It is safe to do, only if the coroutine has finished execution (at the final suspend point) or it is in a suspended state (waiting for some operation to complete), but, somehow, we know that the coroutine will never be resumed by the entity which was supposed to resume the coroutine on behalf of the operation that coroutine is awaiting upon. That is only possible if the underlying asynchronous facility support cancellation.

We strongly believe that support for cancellation is a required facility for writing asynchronous code and we struggled for awhile trying to decide what is the source of the cancellation, whether it is the \tcode{lazy} that must initiate cancellation - and therefore every await in every coroutine needs to understand how to cancel a particular operation it is being awaited upon - or every async operation is tied to a particular lifetime and cancellation domain and operations are cancelled in bulk by cancellation of the entire cancellation domain~\cite{p0399r0}.

We experimented with both approaches and reached the conclusion that not performing cancellation from the \tcode{lazy}, but, pushing it to the cancellation domain leads to more efficient implementation and is a simpler model for users.






\chapter{Discussion}

The following largely omits discussion of the constant variations of the range access CPOs
\tcode{cbegin}, \tcode{cend}, \tcode{crbegin}, \tcode{crend}, and \tcode{cdata}
for brevity despite that the arguments presented for each CPO apply similarly
to its corresponding constant variation.
Imagine that occurrences of e.g. ``\tcode{begin}'' below
are instead occurrences of ``\tcode{begin} (and \tcode{cbegin}).''

\section{arrays of unknown bound}
GB275 states (\cite{GB275}):
\begin{quote}
\tcode{ranges::begin} and \tcode{ranges::end} should not accept arrays of unknown bound.
The current definitions of \tcode{ranges::begin} and \tcode{ranges::end}
mean that an array of unknown bound is treated as an empty range.
The expressions \tcode{E + 0} and \tcode{E + extent_v<T>} are both well-formed
for [an expression \tcode{E} with type \tcode{T} which is] an array of unknown bound
(with \tcode{extent_v<T>} equal to zero).

Proposed change: \\
Make \tcode{ranges::begin(E)} and \tcode{ranges::end(E)} ill-formed
when \tcode{E} is an array of unknown bound.
\end{quote}

In private communication before Belfast,
the author and Jonathan Wakely discussed arrays of unknown bound
and their interactions with the range access CPOs extensively
and came to the conclusion that
\tcode{begin} and \tcode{data} should accept them
to preserve the reasonable behavior of \tcode{std::begin} and \tcode{std::data},
but \tcode{end} and \tcode{size} should not.
It seems logical to relax the \libconcept{range} constraint on \tcode{ranges::iterator_t} as well,
so it can continue to express ``the type that \tcode{ranges::begin} returns''
even in this non-\libconcept{range} case.

\tcode{empty} is also easily supported
despite the lack of extent
since arrays cannot have zero extent.


\section{arrays of elements with incomplete type}

The author and Mr. Wakely also discussed
how the range access CPOs should handle
arrays whose elements have incomplete type.
Such array types were never considered during the evolution of Ranges,
so any status quo behavior is unintended.

\tcode{size} and \tcode{empty} make sense for these arrays since we can determine the extent.
\tcode{begin} and \tcode{data} could be made to work,
but would nonetheless likely produce problems elsewhere in the program since
pointers-to-incomplete-type fail to satisfy \libconcept{weakly_incrementable}
and therefore all of the iterator concepts including notably \libconcept{contiguous_iterator}.
Supporting arrays of incomplete type in the other CPOs seems like a recipe for disaster:
pointer arithmetic doesn't work, and given that the element type may be completed, we're skirting the ODR.
Factoring in the instability around memoization of concepts (See GB046 \cite{GB046}),
we should avoid attempting to support these types in other CPOs for now.


\section{\libconcept{safe_range}s and \tcode{ranges::data}}

\tcode{ranges::begin} avoids returning dangling iterators
with the ``lvalue or \libconcept{safe_range}'' test,
but \tcode{ranges::data} happily returns dangling pointers for rvalue non-\libconcept{safe_range}s.
There's no particular reason for this, other than the fact that
P0970R1 ``Better, Safer Range Access Customization Points'' \cite{p0970r1} and
P0944 ``Contiguous Ranges'' \cite{p0944r0}
were processed at the same time and never reconciled.
This seems like an oversight we can easily correct now
by requiring ``lvalue or \libconcept{safe_range}'' arguments
in \tcode{ranges::data} just as we already do for \tcode{ranges::begin}.


\section{LWG-3258 and poison pills}

LWG-3258 ``Range access and \tcode{initializer_list}'' \cite{lwg3258}
proposes a change to the \tcode{initializer_list} poison pills
for \tcode{begin} and \tcode{end}, and
the addition of similar poison pills to \tcode{rbegin} and \tcode{rend}.
This was a great idea, until
P1870R1 ``\tcode{\placeholdernc{forwarding-range}<T>} is too subtle'' \cite{p1870r1},
changed the opt-in mechanism for \tcode{\placeholdernc{forwarding-range}<T>}
(now \libconcept{safe_range}) from ``do rvalues work?''
to ``is this trait specialized?''.
We could and should have removed the \tcode{initializer_list} poison pills
in P1870R1, but did not - so let's do so now.


\section{P1870 and the advantages of lvalue dispatch}

P1870R1 also modified the design of
\tcode{begin}, \tcode{end}, \tcode{rbegin}, and \tcode{rend}:
these CPOs now only perform lookup and dispatch
with lvalues bound to their argument expression.
For example, \tcode{std::ranges::begin(std::span\{some_array_argument\})} binds
a reference we'll call \tcode{t} to the result of materializing
the prvalue \tcode{std::span\{some_array_argument\}},
determines that \tcode{t.begin()} is a valid expression
whose decayed type models \tcode{input_or_output_iterator},
and then returns \tcode{t.begin()}.
Always performing lookup and dispatch with lvalues in this fashion
makes it easier to reason about, implement, and specify the CPOs.
We should specify the others similarly.

While implementing this change,
I realized that the forwarding-reference poison pills in the working draft
are insufficiently poisonous.
\tcode{void foo(auto\&\&)} is less-specialized than either
\tcode{void foo(auto\&)} or \tcode{void foo(const auto\&)},
so a \tcode{void foo(auto\&\&)} poison pill fails to intercept/ambiguate
calls to such overgeneric lvalue functions as intended.
We should fix the poison pills by replacing them with two lvalue overloads.
(I'm not certain the poison pills serve a useful design purpose anymore,
and I'd like to remove them,
but it's too late in the cycle for even so small a design change.)


\section{ADL only for class/enumeration types}

Just as LWG-3299 doesn't want users to specify non-pointer iterator behaviors
for pointer-to-program-defined-type (\cite{lwg3299}),
we don't intend for users to specify
non-array range behaviors for array-of-program-defined-type.
It's similarly not intended that users specify range behaviors for functions.
We should forbid such silliness in the range access CPOS
just as LWG-3299 does for the iterator machinery
by constraining ADL cases to expressions of class or enumeration type.


\section{Editorial Feedback}
During the merge of P1870R1, the Project Editor expressed dissatisfaction
with a couple of phrases used widely in the CPO wording:
\begin{itemize}
\item ``an lvalue \tcode{t} that denotes the same object as [subexpression] \tcode{E}''
  doesn't make sense when \tcode{E} is a prvalue since prvalues don't ``denote an object.''
  We should instead say that \tcode{T} denotes the result of applying
  the temporary materialization conversion to \tcode{E} when \tcode{E} is a prvalue.
\item ``ranges::begin(E) is ill-formed'' is not something
  to which [an expression] can be expression-equivalent.
  This category error is repeated in every CPO specification
  (sometimes twice after application of P1870R1).
\end{itemize}


\chapter{Implementation Experience}
The proposed changes have been implemented in Microsoft's STL (See https://github.com/microsoft/STL/pull/432).


\chapter{Technical Specifications}
The technical specifications that follow take the form of excerpts
from the working draft with change markings:
\begin{itemize}
\item \removed{Text to be struck is in red with strikethough}, and
\item \added{text to be added is ``green'' with underline}.
\end{itemize}

Note that these specifications supersede the proposed resolution of LWG-3258
and include the proposed resolution of
LWG-3368 ``Exactly when does \tcode{size} return \tcode{end - begin}?'' \cite{lwg3368}.

\setcounter{chapter}{23}
\rSec0[ranges]{Ranges library}
[...]
\setcounter{section}{1}
\rSec1[ranges.syn]{Header \tcode{<ranges>} synopsis}
[...]
\begin{codeblock}
  template<@\changed{range R}{class T}@>
    using iterator_t = decltype(ranges::begin(declval<@\changed{R}{T}@&>()));
\end{codeblock}
[...]
\rSec1[range.access]{Range access}
[...]
\rSec2[range.access.begin]{\tcode{ranges::begin}}
\pnum
The name \tcode{ranges::begin} denotes a customization point
object\cxxiref{customization.point.object}.

\pnum
Given a subexpression \tcode{E} \changed{and an lvalue}{with type \tcode{T}, let}
\tcode{t} \added{be an lvalue} that denotes the same object as \tcode{E}\removed{,}
\added{if \tcode{E} is a glvalue and otherwise denotes the result of applying
the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \changed{if}{If} \tcode{E} is an rvalue and
  \tcode{enable_safe_range<\changed{remove_cvref_t<decltype((E))>}{remove_cv_t<T>}>} is \tcode{false},
  \tcode{ranges::begin(E)} is ill-formed.
  \removed{Otherwise, \tcode{ranges::begin(E)} is expression-equivalent to:}

\item
  \added{Otherwise, if \tcode{T} is an array type\cxxiref{basic.compound} and
  }\tcode{\added{remove_all_extents_t<T>}}\added{ is an incomplete type,
  \tcode{ranges::begin(E)} is ill-formed with no diagnostic required.}

\item
  \changed{\tcode{t + 0} if \tcode{t} is of}{Otherwise, if \tcode{T} is an} array type\removed{\cxxiref{basic.compound}}\added{,
  \tcode{ranges::begin(E)} is expression-equivalent to \tcode{t + 0}}.

\item
  Otherwise, \added{if} \tcode{\placeholdernc{decay-copy}(t.begin())} \removed{if it}
  is a valid expression \changed{and its}{whose} type \removed{\tcode{I}}
  models \libconcept{input_or_output_iterator}\added{,
  \tcode{ranges::begin(E)} is expression-equivalent
  to \tcode{\placeholdernc{decay-copy}(t.begin())}}.

\item
  Otherwise, \added{if \tcode{T} is a class or enumeration type and}
  \tcode{\placeholdernc{decay-copy}(begin(t))} \removed{if it} is a
  valid expression \changed{and its}{whose} type \removed{\tcode{I}} models
  \libconcept{input_or_output_iterator} with overload
  resolution performed in a context \changed{that includes}{in which unqualified lookup for \tcode{begin} finds only} the declarations\removed{:}
\begin{codeblock}
@\removed{template<class T> void begin(T\&\&) = delete;}@
@\removed{template<class T> void begin(initializer_list<T>\&\&) = delete;}@
@\added{void begin(auto\&) = delete;}@
@\added{void begin(const auto\&) = delete;}@
\end{codeblock}
  \removed{and does not include a declaration of \tcode{ranges::begin}}
  \added{then \tcode{ranges::begin(E)} is expression-equivalent
  to \tcode{\placeholdernc{decay-copy}(begin(t))} with overload resolution performed in the above context}.

\item
  Otherwise, \tcode{ranges::begin(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above} result in
substitution failure when \tcode{ranges::begin(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::begin(E)} is a valid expression, its type models
\tcode{input_or_output_iterator}.
\end{note}

\rSec2[range.access.end]{\tcode{ranges::end}}
\pnum
The name \tcode{ranges::end} denotes a customization point
object\cxxiref{customization.point.object}.

\pnum
Given a subexpression \tcode{E} \changed{and an lvalue}{with type \tcode{T}, let}
\tcode{t} \added{be an lvalue} that denotes the same object as \tcode{E}\removed{,}
\added{if \tcode{E} is a glvalue and otherwise denotes the result of applying
the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \changed{if}{If} \tcode{E} is an rvalue and
  \tcode{enable_safe_range<\changed{remove_cvref_t<decltype((E))>}{remove_cv_t<T>}>} is \tcode{false},
  \tcode{ranges::end(E)} is ill-formed.
  \removed{Otherwise, \tcode{ranges::end(E)} is expression-equivalent to:}

\item
  \added{Otherwise, if \tcode{T} is an array type\cxxiref{basic.compound} and
  }\tcode{\added{remove_all_extents_t<T>}}\added{ is an incomplete type,
  \tcode{ranges::end(E)} is ill-formed with no diagnostic required.}

\item
  \added{Otherwise, if \tcode{T} is an array of unknown bound,
  \tcode{ranges::end(E)} is ill-formed.}

\item
  \added{Otherwise, if \tcode{T} is an array, \tcode{ranges::end(E)} is expression-equivalent to}
  \tcode{t + extent_v<T>} \removed{if \tcode{E} is of array type\cxxiref{basic.compound} \tcode{T}}.

\item
  Otherwise, \added{if}
  \tcode{\placeholdernc{decay-copy}(t.end())}
  \removed{if it} is a valid expression \changed{and its}{whose} type \removed{\tcode{S}} models
\begin{codeblock}
sentinel_for<@\changed{decltype(ranges::begin(E))}{iterator_t<T>}@>
\end{codeblock}
  \added{then \tcode{ranges::end(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(t.end())}.}

\item
  Otherwise, \removed{\tcode{\placeholdernc{decay-copy}(end(t))}} if \removed{it}
  \added{\tcode{T} is a class or enumeration type and
  \tcode{\placeholdernc{decay-copy}(end(t))}} is a valid
  expression \changed{and its}{whose} type \removed{\tcode{S}} models
\begin{codeblock}
sentinel_for<@\changed{decltype(ranges::begin(E))}{iterator_t<T>}@>
\end{codeblock}
  with overload
  resolution performed in a context \changed{that includes}{in which unqualified lookup for \tcode{end} finds only} the declarations\removed{:}
\begin{codeblock}
@\removed{template<class T> void end(T\&\&) = delete;}@
@\removed{template<class T> void end(initializer_list<T>\&\&) = delete;}@
@\added{void end(auto\&) = delete;}@
@\added{void end(const auto\&) = delete;}@
\end{codeblock}
  \removed{and does not include a declaration of \tcode{ranges::end}}
  \added{then \tcode{ranges::end(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(end(t))} with overload resolution performed in the above context}.

\item
  Otherwise, \tcode{ranges::end(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above}
result in substitution failure when \tcode{ranges::end(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::end(E)} is a valid expression,
the types \tcode{S} and \tcode{I} of
\tcode{ranges::end(E)} and \tcode{ranges::begin(E)}
model \tcode{\libconcept{sentinel_for}<S, I>}.
\end{note}

[...]

\setcounter{subsection}{4}
\rSec2[range.access.rbegin]{\tcode{ranges::rbegin}}
\pnum
The name \tcode{ranges::rbegin} denotes a customization point
object\cxxiref{customization.point.object}.

\pnum
Given a subexpression \tcode{E} \changed{and an lvalue}{with type \tcode{T}, let}
\tcode{t} \added{be an lvalue} that denotes the same object as \tcode{E}\removed{,}
\added{if \tcode{E} is a glvalue and otherwise denotes the result of applying
the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \changed{if}{If} \tcode{E} is an rvalue and
  \tcode{enable_safe_range<\changed{remove_cvref_t<decltype((E))>}{remove_cv_t<T>}>} is \tcode{false},
  \tcode{ranges::rbegin(E)} is ill-formed.
  \removed{Otherwise, \tcode{ranges::rbegin(E)} is expression-equivalent to:}

\item
  \added{Otherwise, if \tcode{T} is an array type\cxxiref{basic.compound} and
  }\tcode{\added{remove_all_extents_t<T>}}\added{ is an incomplete type,
  \tcode{ranges::rbegin(E)} is ill-formed with no diagnostic required.}

\item
  \added{Otherwise, if} \tcode{\placeholdernc{decay-copy}(t.rbegin())}
  \removed{if it} is a valid expression \changed{and its}{whose} type \removed{\tcode{I}} models
  \libconcept{input_or_output_iterator}\added{,
  \tcode{ranges::rbegin(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(t.rbegin())}}.

\item
  Otherwise, \removed{\tcode{\placeholdernc{decay-copy}(rbegin(t))}} if \removed{it}
  \added{\tcode{T} is a class or enumeration type and
  \tcode{\placeholdernc{decay-copy}(rbegin(t))}} is a valid
  expression \changed{and its}{whose} type \removed{\tcode{I}} models
  \libconcept{input_or_output_iterator} with overload
  resolution performed in a context \changed{that includes}{in which unqualified lookup for \tcode{rbegin} finds only} the declaration\changed{:}{s}
\begin{codeblock}
@\removed{template<class T> void rbegin(T\&\&) = delete;}@
@\added{void rbegin(auto\&) = delete;}@
@\added{void rbegin(const auto\&) = delete;}@
\end{codeblock}
  \removed{and does not include a declaration of \tcode{ranges::rbegin}}
  \added{then \tcode{ranges::rbegin(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(rbegin(t))} with overload resolution performed in the above context}.

\item
  Otherwise, \tcode{\removed{make_reverse_iterator(ranges::end(t))}} if both
  \tcode{ranges::begin(t)} and \tcode{ranges::end(t)} are valid
  expressions of the same type \removed{\tcode{I}} which models
  \libconcept{bidirectional_iterator}\cxxiref{iterator.concept.bidir}\added{,
  \tcode{ranges::rbegin(E)} is expression-equivalent to }\tcode{\added{make_reverse_iterator(ranges::end(t))}}.

\item
  Otherwise, \tcode{ranges::rbegin(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above}
 result in substitution failure when \tcode{ranges::rbegin(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::rbegin(E)} is a valid expression, its type models
\libconcept{input_or_output_iterator}.
\end{note}

\rSec2[range.access.rend]{\tcode{ranges::rend}}
\pnum
The name \tcode{ranges::rend} denotes a customization point
object\cxxiref{customization.point.object}.

\pnum
Given a subexpression \tcode{E} \changed{and an lvalue}{with type \tcode{T}, let}
\tcode{t} \added{be an lvalue} that denotes the same object as \tcode{E}\removed{,}
\added{if \tcode{E} is a glvalue and otherwise denotes the result of applying
the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \changed{if}{If} \tcode{E} is an rvalue and
  \tcode{enable_safe_range<\changed{remove_cvref_t<decltype((E))>}{remove_cv_t<T>}>} is \tcode{false},
  \tcode{ranges::rend(E)} is ill-formed.
  \removed{Otherwise, \tcode{ranges::rend(E)} is expression-equivalent to:}

\item
  \added{Otherwise, if \tcode{T} is an array type\cxxiref{basic.compound} and
  }\tcode{\added{remove_all_extents_t<T>}}\added{ is an incomplete type,
  \tcode{ranges::rend(E)} is ill-formed with no diagnostic required.}

\item
  \added{Otherwise, if} \tcode{\placeholdernc{decay-copy}(t.rend())}
  \removed{if it} is a valid expression \changed{and its}{whose} type \removed{\tcode{S}} models
\begin{codeblock}
sentinel_for<decltype(ranges::rbegin(E))>
\end{codeblock}
  \added{then \tcode{ranges::rend(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(t.rend())}.}

\item
  Otherwise, \removed{\tcode{\placeholdernc{decay-copy}(rend(t))}} if \removed{it}
  \added{\tcode{T} is a class or enumeration type and
  \tcode{\placeholdernc{decay-copy}(rend(t))}} is a valid
  expression \changed{and its}{whose} type \removed{\tcode{S}} models
\begin{codeblock}
sentinel_for<decltype(ranges::rbegin(E))>
\end{codeblock}
  with overload
  resolution performed in a context \changed{that includes}{in which unqualified lookup for \tcode{rend} finds only} the declaration\changed{:}{s}
\begin{codeblock}
@\removed{template<class T> void rend(T\&\&) = delete;}@
@\added{void rend(auto\&) = delete;}@
@\added{void rend(const auto\&) = delete;}@
\end{codeblock}
  \removed{and does not include a declaration of \tcode{ranges::rend}}
  \added{then \tcode{ranges::rbegin(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(rend(t))} with overload resolution performed in the above context}.

\item
  Otherwise, \tcode{\removed{make_reverse_iterator(ranges::begin(t))}} if both
  \tcode{ranges::begin(t)} and \tcode{ranges::end(t)} are valid
  expressions of the same type \removed{\tcode{I}} which models
  \libconcept{bidirectional_iterator}\cxxiref{iterator.concept.bidir}\added{,
  then \tcode{ranges::rend(E)} is expression-equivalent to }\tcode{\added{make_reverse_iterator(ranges::begin(t))}}.

\item
  Otherwise, \tcode{ranges::rend(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above}
result in substitution failure when \tcode{ranges::rend(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::rend(E)} is a valid expression,
the types \tcode{S} and \tcode{I} of
\tcode{ranges::rend(E)} and \tcode{ranges::rbegin(E)}
model \tcode{\libconcept{sentinel_for}<S, I>}.
\end{note}

[...]
\setcounter{subsection}{8}
\rSec2[range.prim.size]{\tcode{ranges::size}}
\pnum
The name \tcode{\added{ranges::}size} denotes a customization point
object\cxxiref{customization.point.object}.

\pnum
\changed{The expression \tcode{ranges::size(E)} for some}{Given a} subexpression \tcode{E} with type
\tcode{T} \changed{is expression-equivalent to:}{, let \tcode{t} be an lvalue that denotes the same object as \tcode{E} if \tcode{E} is a glvalue and otherwise denotes the result of applying the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \added{If \tcode{T} is an array of unknown bound\cxxiref{dcl.array},
  \tcode{ranges::size(E)} is ill-formed.}

\item
  \tcode{\removed{\placeholdernc{decay-copy}(extent_v<T>)}} \added{Otherwise,} if \tcode{T} is an array
  type\removed{\cxxiref{basic.compound}}\added{,
  \tcode{ranges::size(E)} is expression-equivalent to }\tcode{\added{\placeholdernc{decay-copy}(extent_v<T>)}}.

\item
  \removed{Otherwise, if}
  \tcode{\removed{disable_sized_range<remove_cv_t<T>>}}\removed{\cxxiref{range.sized}
  is \tcode{false}:}

\item
  \added{Otherwise,} \removed{\tcode{\placeholdernc{decay-copy}(E.size())}}
  if \removed{it} \tcode{\added{disable_sized_range<remove_cv_t<T>>}}\added{\cxxiref{range.sized}
  is \tcode{false} and \tcode{\placeholdernc{decay-copy}(t.size())}}
  is a valid expression \changed{and its}{of integer-like} type\removed{ \tcode{I}
  is integer-like}\cxxiref{iterator.concept.winc}\added{,
  \tcode{ranges::size(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(E.size())}}.

\item
  Otherwise, \removed{\tcode{\placeholdernc{decay-copy}(size(E))}} if \removed{it}
  \added{\tcode{T} is a class or enumeration type,
  }\tcode{\added{disable_sized_range<remove_cv_t<T>>}}\added{ is \tcode{false},
  and }\tcode{\added{\placeholdernc{decay-copy}(size(t))}} is
  a valid expression \changed{and its}{of integer-like} type \removed{\tcode{I}
  is integer-like} with overload resolution performed in a context
  \changed{that includes}{in which unqualified lookup for \tcode{size} finds only}
  the declaration\changed{:}{s}
\begin{codeblock}
@\removed{template<class T> void size(T\&\&) = delete;}@
@\added{void size(auto\&) = delete;}@
@\added{void size(const auto\&) = delete;}@
\end{codeblock}
  \removed{and does not include a declaration of \tcode{ranges::size}}
  \added{then \tcode{ranges::size(E)} is expression-equivalent
  to \tcode{\placeholdernc{decay-copy}(size(E))} with overload resolution performed in the above context}.

\item
  Otherwise, \added{if} \tcode{\placeholdernc{make-unsigned-like}(ranges::end(\changed{E}{t}) - ranges::begin(\changed{E}{t}))}~(\changed{\cxxref{range.subrange}}{\cxxref{range.syn}})
  \removed{if it} is a valid expression and
  the types \tcode{I} and \tcode{S} of \tcode{ranges::begin(\changed{E}{t})} and
  \tcode{ranges::end(\changed{E}{t})} (respectively) model both
  \tcode{\libconcept{sized_sentinel_for}<S, I>}\cxxiref{iterator.concept.sizedsentinel} and
  \tcode{\libconcept{forward_iterator}<I>}\added{,
  then \tcode{ranges::size(E)} is expression-equivalent to
  }\tcode{\added{\placeholdernc{make-unsigned-like}(ranges::end(t) - ranges::begin(t))}}.
  \removed{However, \tcode{E} is evaluated only once.}

\item
  Otherwise, \tcode{ranges::size(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above}
result in substitution failure when \tcode{ranges::size(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::size(E)} is a valid expression, its
type is integer-like.
\end{note}

\rSec2[range.prim.empty]{\tcode{ranges::empty}}
\pnum
The name \tcode{\added{ranges::}empty} denotes a customization point
object\cxxiref{customization.point.object}.

\pnum
\changed{The expression \tcode{ranges::empty(E)} for some}{Given a} subexpression \tcode{E}
\changed{is expression-equivalent to:}{with type \tcode{T}, let \tcode{t} be an lvalue that denotes the same object as \tcode{E} if \tcode{E} is a glvalue and otherwise denotes the result of applying the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \added{If \tcode{T} is an array of unknown bound\cxxiref{basic.compound},
  \tcode{ranges::empty(E)} is ill-formed.}

\item
  \added{Otherwise, if} \tcode{bool(\changed{(E)}{t}.empty())} \removed{if it} is a valid expression\added{,
  \tcode{ranges::empty(E)} is expression-equivalent to \tcode{bool(t.empty())}}.

\item
  Otherwise, \added{if} \tcode{(ranges::size(\changed{E}{t}) == 0)} \removed{if it} is a valid expression\added{,
  \tcode{ranges::empty(E)} is expression-equivalent to \tcode{(ranges::size(t) == 0)}}.

\item
  Otherwise, \changed{\tcode{EQ}, where \tcode{EQ} is}{if}
  \tcode{bool(ranges::begin(\changed{E}{t}) == ranges::end(\changed{E}{t}))}
  \removed{except that \tcode{E} is only evaluated once,}
  \removed{if \tcode{EQ}} is a valid expression and
  the type of \tcode{ranges::begin(\changed{E}{t})} models \libconcept{forward_iterator}\added{,
  \tcode{ranges::empty(E)} is expression-equivalent to \tcode{bool(ranges::begin(t) == ranges::end(t))}}.

\item
  Otherwise, \tcode{ranges::empty(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above}
result in substitution failure when \tcode{ranges::empty(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::empty(E)} is a valid expression,
it has type \tcode{bool}.
\end{note}

\rSec2[range.prim.data]{\tcode{ranges::data}}
\pnum
The name \tcode{\added{ranges::}data} denotes a customization point object\cxxiref{customization.point.object}.

\pnum
\changed{The expression \tcode{ranges::data(E)} for some}{Given a} subexpression \tcode{E}
\changed{is expression-equivalent to:}{with type \tcode{T},
let \tcode{t} be an lvalue that denotes the same object as \tcode{E} if \tcode{E} is a glvalue and otherwise denotes the result of applying the temporary materialization conversion\cxxiref{conv.rval} to \tcode{E}. Then:}

\begin{itemize}
\item
  \added{If \tcode{E} is an rvalue and
  }\tcode{\added{enable_safe_range<remove_cv_t<T>>}}\added{ is \tcode{false},
  \tcode{ranges::data(E)} is ill-formed.}

\item
  \added{Otherwise, if \tcode{T} is an array type\cxxiref{basic.compound} and
  }\tcode{\added{remove_all_extents_t<T>}}\added{ is an incomplete type,
  \tcode{ranges::data(E)} is ill-formed with no diagnostic required.}

\item
  \changed{If \tcode{E} is an lvalue,}{Otherwise, if}
  \tcode{\placeholdernc{decay-copy}(\changed{E}{t}.data())}
  \removed{if it} is a valid expression of pointer to object type\added{,
  \tcode{ranges::data(E)} is expression-equivalent to \tcode{\placeholdernc{decay-copy}(t.data())}}.

\item
  Otherwise, if \tcode{ranges::begin(\changed{E}{t})} is a valid expression
  whose type models \libconcept{contiguous_iterator},
  \added{\tcode{ranges::data(E)} is expresssion-equivalent to} \tcode{to_address(ranges::begin(E))}.

\item
  Otherwise, \tcode{ranges::data(E)} is ill-formed.
\end{itemize}

\pnum
\begin{note}
\changed{This case can}{Diagnosable ill-formed cases above}
result in substitution failure when \tcode{ranges::data(E)}
appears in the immediate context of a template instantiation.
\end{note}

\pnum
\begin{note}
Whenever \tcode{ranges::data(E)} is a valid expression, it
has pointer to object type.
\end{note}
